%
% Acknowledgements
%

\chapter{Acknowledgements}

\enquote{The more you know, the more you know you don't know}.
No other aphorism may be more apt at describing my perspective on these past four years,
but what I do know---without a doubt---is that I owe a great deal to a great many people.

First and foremost, I must thank my supervisor, Carl Anderson.
I greatly admire your commitment to embodying a culture of rigorous first-class science,
and equally your commitment to the prosperity and well-being of everyone under your wing. 
I am indebted for all of your support and guidance, 
and your empathy and optimism during interesting times.
Thank you for the privilege of the past four years---it has been an honour and a pleasure.

I must also take the opportunity to address the members of the Anderson team, past and present.
% iterate through team members past and present
The team has greatly evolved these past few years both in scientific direction and in membership,
but it has always remained a friendly and enriching place to be. 
Whether over the lunch table in Murrays,
during our legendary team retreats,
or in our traditionally lively team meetings,
there was no lack of supportive and productive discussion.
In particular, I must thank
    Aleksejs Sazonovs,
    Carla Jones-Bell,
    Elizabeth Goode,
    Laura Fachal,
    Leland Taylor,
    Loukas Moutsianas, 
    Nikolaos Panousis,
    and Velislava Petrova,
    each of whom lent their ear to me countless times.

The research presented in this thesis would not have been possible without the help of numerous individuals and organisations.
The projects in this thesis are highly collaborative, extending from the Wellcome Sanger Institute to London, Exeter, and the US.
To Adrian Hayday, Nicholas Kennedy, and Tariq Ahmad, 
thank you for granting me the opportunity to be part of these amazing collaborations.
I owe a sincere gratitude to the many clinicians, scientists, and administrators who laid the foundation for this thesis,
and even more so to the patients who contributed their samples.
I would also like to thank the Wellcome Trust and other funding agencies that contributed to the projects in my thesis for their generous financial support.

Thank you to my thesis committee, Daniel Gaffney and Michael Inouye, who provided valuable feedback throughout the course of my PhD.
I also extend my gratitude to the amazing research administrators I've worked with at Sanger: Carol Dunbar, Sally Bygraves, Eloise Stapleton, Paris Litterick, Rachel Henry, and Sophie Leggett.
I'm sure there has been many a time you've known more about my projects than me. 
% Without you all my research would have struggled to stay afloat.
To the Sanger Sample Management, Pipelines, and Informatics teams:
the incredible work you do to keep the sequencing and computing resources running smoothly was critical for this thesis, and for all science done on the Campus.

On the other side of the balance, I would like thank all the friends I have made during my time in Cambridge.
To my fellow PhD cohort, 
I'll remember all the punting trips, the pub quizzes, the eSCAMPS dinners,
and our shared commiserations on the Campus bus.
To the Churchill College badminton team: 
thank you for having me, and best wishes for all the training sessions and league matches in the years to come.
I would like to give a special mention to my friends from the Cambridge University Anime and Manga Society.
Many good times were had; many good memories were made.
You've truly made my \enquote{campus life} very much multicoloured.

Finally, to my family; to my brother, my mother, my father, my grandparents:
I could always depend on your unconditional love and support.
Despite the thousands of miles between us, 
all of you were held close to my heart.
%
